Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-03-11 draft.tex
.

\usepackage[titletoc]{appendix}

\captionsetup[table]{singlelinecheck=false}
\captionsetup[figure]{singlelinecheck=false}
\author{Fabian Schubert}

@@

\tableofcontents
\newpage
\clearpage
\setcounter{page}{1}

@@

Theoretical models of homeostasis are generally based on a form of feedback control. This is also known as a \textit{closed loop control} in control theory. All these systems include a number of necessary functional components, which will be briefly introduced in the following. Concepts and terms were taken from \cite[p. 59-64]{Cybernetik_Systems_Cruse_2006}.
\newpage
\subsubsection{Theory of Control Systems} \label{Control_Sys_Theo_Section}
A closed loop control generally must contain some \emph{process} that is to be regulated. Despite its potential internal complexity, its functional role in the context of a feedback system is to simply generate a time-dependent output based on the history of a time-dependent input. This output is what one wishes to control and regulate by means of the feedback loop and thus represents the signal that is further processed. It might not be possible however to directly quantify this desired output in terms of some internal state of the process. The firing rate of a neuron for example is an abstract quantity, which can not be \emph{directly} related to some neuronal state variable. Consequently, it might be necessary to transform the actual output of the process into a different signal by a \textit{feedback transducer}. The (potentially transformed) output is fed into a \textit{comparator} in the next step, which---unsurprisingly---compares the given output of the process to some given target or \textit{reference}, typically by subtraction. Note that this reference can but does note have to be a time-dependent signal. The term \textit{homeostasis} however generally refers to a fixed reference, while implementations with a time dependent reference can be regarded as a form of \textit{servomechanism}. The difference between reference and the feedback transducer's signal is known as the \textit{error} signal. This error signal is given to a \textit{controller}, which transforms this error signal into an appropriate input signal for the initially introduced process. This is also the point in the control loop where \textit{external disturbances} come into play by being added to the input generated by the controller. We illustrated the described flow of feedback control in Figure \ref{Closed_Loop_Control_Illustration}.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/illustrations/closed_loop_control.eps}
\end{center}
\caption{Schematic illustration of a closed loop control as described in Section \ref{Control_Sys_Theo_Section}}
\label{Closed_Loop_Control_Illustration}
\end{figure}
Based on the way the controller generates input for the dynamic process based on the error signal one can classify control loops by three types:
\newpage
\begin{itemize}
\item A \textit{proportional controller} generates a signal that is proportional to the error signal it receives. The most important property of a proportional controller is that the stationary state of the control loop generally does not imply an error signal equal to zero.
\item An \textit{integrator controller} uses an integration over the past of the error signal to obtain its output. The stationary solution necessarily implies that the error signal is zero.
\item A \textit{differential controller} generates its output by means of the time derivative of the error signal. This type of controller can quickly react to momentary changes but does not compensate for sustained deviations of the error signal from zero.
\end{itemize}
It should be pointed out that these types of control mechanisms can be simultaneously used in a control system, e.g. through a linear combination. Furthermore, the presented terms refer to concepts of linear control theory. In the study of nonlinear feedback systems they thus only apply in a limited regime where a first order approximation of the system is sufficiently exact.

@@

\begin{table}
\begin{tabu}{|l|l|l|l|l|}
\hline
\textbf{parameter} & \textbf{EE} & \textbf{EI} & \textbf{IE} & \textbf{II} \\ \hline
connection fraction & $\rightarrow 0.1$ & $0.1$ & $0.1$ & $0.5$ \\ \hline
initial connection strength & $0.0001 mV$ & $1.5 mV$ & $-1.5 mV$ & $-1.5 mV$ \\ \hline
conduction delay & $1.5 ms$ & $0.5 ms$ & $1.0 ms$ & $1.0 ms$ \\
\hline
\end{tabu}
\caption{Parameters of synaptic connections.}
\label{syn_conn_params}
\end{table}
\newpage

\subsubsection{Neuron Model}
We used a leaky integrate-and-fire model for all neurons in the network, whose dynamics are described by a stochastic differential equation:
\begin{equation}
{\tau_m}dV_i = -\left( V_i-E_l \right) dt + \sqrt{\tau_m} \sigma dW_i + \tau_m \sum_{j,k} w^{\rm effective}_{ij}\delta \left( t^k_{\mathrm{spike},j} + t_{\rm delay} - t \right)
\label{LIF_Dynamics}
\end{equation}
where $V_i$ is the membrane potential of neuron $i$, $E_l$ is the equilibrium membrane potential, $\tau_m$ is the time constant of the membrane, $\sigma$ is the standard deviation of the noise term and $dW$ is the standard Wiener process. A neuron is said to spike when its membrane potential reaches the threshold voltage $V_t$. The voltage is then reset to $V_r$. A refractory period was not implemented. A presynaptic spike from neuron $j$ causes a simple (delayed, see Table \ref{syn_conn_params}) increment of the membrane potential of the postsynaptic neuron $i$ by $w^{\rm effective}_{ij}$. Table \ref{LIF_neuron_params} summarizes the aforementioned set of parameters.
\begin{table}[H]
\begin{tabu}{|l|l|l|}
\hline
\textbf{parameter} & \textbf{exc. neur.} & \textbf{inh. neur.}\\ \hline
$\mathrm{E_l}$ & $\mathrm{-60\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\tau_m}$ & $\mathrm{20\;ms}$ & $\mathrm{20\;ms}$ \\ \hline
$\mathrm{V_r}$ & $\mathrm{-70\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\sigma}$ & $\mathrm{\sqrt{5}\;mV}$ & $\mathrm{\sqrt{5}\;mV}$ \\ \hline
$\mathrm{V_t}$ & subject to IP & $\mathrm{-58\;mV}$ \\ 
\hline
\end{tabu}
\caption{Parameters of LIF neuron}
\label{LIF_neuron_params}
\end{table}

@@

Finally, the dynamics of firing thresholds $V_{t,i}$ were modeled such that the rate of change is proportional to the relative deviation of $NO$ concentration at the neurons' locations from a global target concentration $NO_0$.

The appropriate choice of $NO_0$ is obviously crucial for the goal of achieving and maintaining a certain level of activity. However, one cannot directly set a parameter of the model to the desired population activity as was the case for canonical intrinsic homeostasis. One rather needs to determine the average concentration \textit{associated} with the desired activity and set it as a target concentration. Though it is possible to derive this relation in an analytic fashion, we let the system run with the previous homeostatic mechanism, still solving equation \eqref{Ca_dyn}--\eqref{NO_dyn} until a steady mean over the concentrations at the neurons' positions was reached. This mean was then set to be the target concentration and we switched to diffusive homeostasis. Table \ref{Params_IP} summarizes the choice of parameters that were introduced in this section. Diffusion parameters roughly match those measured in experiments \cite{Philippides_2000}.
\begin{table}
\begin{tabu}{|l|l|}
\hline
\textbf{parameter} & \textbf{value} \\
\hline
$\mathrm{r_{IP}}$ & $\mathrm{3\;Hz}$ \\
\hline
$\mathrm{\eta_{IP}}$ & $\mathrm{0.1\;mV}$ \\
\hline
$\mathrm{Ca^{2+}_{spike}}$ & 1 \\ \hline
$\mathrm{\tau_{Ca^{2+}}}$ &  $\mathrm{10\;ms}$ \\
\hline
$\mathrm{\tau_{nNOS}}$ & $\mathrm{100\;ms}$ \\
\hline
$\mathrm{D}$ & default: 10 $\mathrm{\mu m^2 ms^{-1}}$ \\
\hline 
$\mathrm{\lambda}$ & $\mathrm{0.1\;s^{-1}}$ \\
\hline
$\mathrm{\tau_{V_t}}$ & see Section \ref{activ_analys} \\
\hline
\end{tabu}
\caption{Parameters of homeostatic intrinsic plasticity.}
\label{Params_IP}
\end{table}

@@

3.) Dirichlet boundary conditions with $NO = NO_{bound}$ at the boundaries.\\
Neumann boundary conditions were used for most of the simulations if not explicitly marked differently. This decision relates to the previously described mechanism of synaptic growth: Neurons placed close to the edge of the sheet have a lower connection probability due to the absence of neighboring neurons in the direction perpendicular to the close-by border. It therefore models the synaptic growth within a square ``cutout" of neural tissue. The Neumann boundary condition fits into this picture since it allows a zero-flux condition at the borders. This is a reasonable assumption, because NO molecules cannot diffuse out of the tissue (unless they were placed in a fluid surrounding).

@@

\subsection{Activity Analysis} \label{activ_analys}
As a first step, we investigated how diffusive homeostasis compares to non-diffusive homeostasis in terms of maintaining a stable level of population activity among excitatory and inhibitory population (though the latter one is only indirectly affected by excitatory synaptic input).
\subsubsection{Dynamic Behavior of the Population Activity}
We set the time constant of threshold adaptation in the diffusive homeostasis to $\mathrm{2500\, ms}$ on a first attempt, as given by \cite{Sweeney_Paper}. Switching between non-diffusive/diffusive homeostasis after $\rm 750\, s$, Figure \ref{full_sim_osci} shows the resulting dynamics of the population activity of the excitatory and inhibitory group (panel A/A*), the average $\rm NO$ concentration at the excitatory neurons' positions (panel B/B*) and the average firing threshold (panel C/C*) within the excitatory group (inhibitory threshold was fixed, see table \ref{LIF_neuron_params}). Both homeostatic mechanisms managed to keep the excitatory population activity in the desired range of 3 Hz. However, what might appear to be slightly faster and stronger random fluctuations in panel A-C were in fact regular oscillations across all three depicted variables, as can be seen in panel A*-C*, showing the dynamics in a smaller time window. While the oscillation amplitudes undergo a rather unpredictable course, the frequency remains at a constant level of $\simeq 0.5\, Hz$. This feature is also illustrated by the fact that the power spectra depicted in Figure \ref{Power_Spec_without_Analysis} have a peak at the frequency that appeared to be dominant in Figure \ref{full_sim_osci}. The fact that the mean over all power spectra of excitatory thresholds differs from the power spectrum of the mean of these thresholds in its amplitude indicates that not all thresholds oscillate at the same phase. The overall shapes of both spectra are still similar, though.
\begin{figure}
\begin{center}
%% oscillating_activity_plot.py
\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_compl.png}
\end{center}
%\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_closeup.png}
\caption{A/A*: Mean of excitatory (blue) and inhibitory (red) neuronal firing rate. B/B*: NO concentration averaged over readouts at excitatory neurons' positions. C/C*: Average excitatory firing threshold. Non-diffusive homeostasis was used for $\mathrm{0-700\, s}$ , diffusive h. for $\mathrm{700-1500\, s}$. Target activity was $\mathrm{3\, Hz}$.}
\label{full_sim_osci}
\end{figure}
\begin{figure}
\begin{center}
%%% plot_pow_spec.py
\includegraphics[width=0.7\textwidth]{../../plots/power_spec/power_spectrum_sim_only.png}
\caption{Blue: Mean of power spectrum over all excitatory thresholds. Red: Power spectrum of mean over all excitatory thresholds.}
\label{Power_Spec_without_Analysis}
\end{center}
\end{figure}

@@

where $\tau_e$ and $\tau_i$ are time constants representing the rapidity of firing rate dynamics, $\phi_e$ and $\phi_i$ the respective transfer functions between synaptic input and firing rate and $W_{xy}$ the mean synaptic weight from population x to population y. A persistent oscillation then corresponds to a stable limit cycle in the two-dimensional phase plane. Such a limit cycle exists inside a region $R$ if $R$ contains no fixed points and if any trajectory whose starting points lies within $R$ is confined to $R$, according to the Poincaré-Bendixson theorem. The second condition is equivalent to the condition that for all points $\mathbf{r}_{edge}$ on the edge of $R$, the vector field $\dot{\mathbf{r}}_{edge}$ is facing into $R$, see \cite[p.~248]{Dyn_Sys_Hirsch}. This case occurs if the system has a locally unstable fixed point, but a non-linearity that prevents an infinite deviation from the fixed point.

@@

\caption{Comparison of mean firing rate of a LIF neuron without noise (squares) and with noise (circles, $\mathrm{\sigma = \sqrt{5}\, mV}$), averaged over $\mathrm{50\, s}$. Red and blue curves are the corresponding analytic predictions, see equation \ref{LIF_Fir_Rate_no_Noise} and \ref{LIF_Fir_Rate_with_Noise1}.}

@@

\sigma_{tot.} = \sqrt{\sigma_{\rm intr.}^2 + \sum_A \sigma_{AB}^2} \; .

@@

x_{st.}(r) &= \frac{1-\exp \left(-\frac{1}{r\cdot \tau _d}\right)}{1-\left(1-u_{st.}\left(r\right)\right)\cdot \exp \left(-\frac{1}{r\cdot \tau _d}\right)} \label{STP_steady1} \\
u_{st.}(r) &= \frac{U}{1-\left(1-U\right)\exp \left(-\frac{1}{r\cdot \tau _f}\right)} \; .\label{STP_steady2}

@@

One thus needs to incorporate the factor $\mathrm{x_{st.}\cdot u_{st.}}$ by means of the previously described mean input and the contribution to the total variance of the membrane noise when trying to find the fixed point of Equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}.

All these aspects taken together make it impossible to find an analytic expression for the fixed point of Equation \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}. Running a simulation of Equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}, including all aforementioned approximations resulted in a stable configuration. We set the relevant parameters to their predefined values for this simulation and additionally extracted the mean threshold of the excitatory population which was set by the non-diffusive homeostasis, see C/C* in Figure \ref{full_sim_osci}, yielding a threshold of $\mathrm{-56.963\, mV}$. The results are depicted in Figure \ref{Dyn_Rate_Approx}. The system settled to a fixed point of $\mathrm{(r_{0,e},r_{0,i}) = (2.992\, Hz, 7.144\, Hz)}$, which is quite close to $\mathrm{(3\, Hz, 6.768\, Hz)}$, being the mean excitatory and inhibitory firing rate found between $\mathrm{100-750\, s}$ of the full simulation (i.e., during non-diffusive homeostasis).
\begin{figure}
%% fixed_point_net.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/approx_rates_EI.png}
\end{center}
\caption{Dynamics of equations \ref{2d-exc_inh_rate_1} (blue line) and \ref{2d-exc_inh_rate_2} (red line). Dotted lines mark the mean frequencies that were present in the full spiking network. Excitatory: 3 Hz, Inhibitory: 6,768 Hz. \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} converged to 2.992 Hz and 7.144 Hz respectively.}
\label{Dyn_Rate_Approx}
\end{figure}
In principle, one could now further quantify the stability of the found fixed point by calculating the Jacobian matrix at $\mathrm{(r_{0,e},r_{0,i})}$. For the sake of the initial question, though, namely the source of oscillation, it is sufficient at this point to restrict oneself to a preliminary result: The recurrent network as such (without homeostasis) is stable under the given choice of parameters, assuming validity of the described rate model. Furthermore, on an empirical basis, the occurrence of oscillations appears to be rather dependent on the choice of homeostasis since all parameters of the simulation were kept unchanged upon the transition after 750 s, apart from the homeostatic mechanism itself. Therefore, we present in the following section our further analysis of the dynamics underlying the diffusive homeostatic feedback.

@@

Both forms of homeostasis use excitatory firing thresholds as a means of adjusting the excitatory firing rate. Though having a relatively immediate impact on the firing rate of the particular neuron, the network as a whole reacts by means of a local disturbance of activity as well: It settles at a new fixed point of firing rates. Two questions are of importance in the context of feedback dynamics. First---obviously---how the relation between a local change of threshold and the new fixed point of the network can be described and second, whether the time scale of the network's response is of relevance with respect to other time scales involved in the feedback loop. Regarding the second question, Figure \ref{rate_th_close} suggests that the excitatory population activity follows the mean firing threshold in a quasi instantaneous fashion, at least in comparison with the overall time scale of the oscillating pattern.
\begin{figure}
%% oscillating_activity_plot_rate_threshold.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/diff_hom/rate_th_close.png}
\end{center}
\caption{Time course of the mean firing rate (blue) and the mean firing threshold within the excitatory pool (red). Bin width for firing rate estimation was 0.1 s.}
\label{rate_th_close}
\end{figure}
We thus presumed an immediate functional relationship $\mathbf{r}_e (t) = \mathbf{r}_e (\mathbf{V}_{t,e} (t))$ in a first attempt at understanding the occurrence of undesired oscillations, where $\mathbf{r}_e$ and $\mathbf{V}_{t,e}$ represent the set of excitatory firing rates and thresholds, respectively.

@@

\caption{Distribution of firing rates for $\mathrm{D=10\, \mu m^2 ms^{-1}}$, full simulation and analytic prediction. Data was taken from 10 simulation runs.}

>>>>

\usepackage[titletoc]{appendix}
\captionsetup[table]{singlelinecheck=false}
\captionsetup[figure]{singlelinecheck=false}

\author{Fabian Schubert}

@@

\tableofcontents
\newpage
\clearpage

\setcounter{page}{1}

\section*{Glossary}
\bigskip
\begin{minipage}{0.33\textwidth} 
$\rm Ca^{2+}$\\
CDF\\
CF\\
CNS\\
CV\\
IP\\
ISI\\
$\rm K^{+}$\\
LIF\\
LTD\\
LTP\\
L5\\
MLE\\
nNOS\\
NO\\
PDF\\
SORN\\
STD\\
STDP\\
STF\\
STP
\end{minipage}
\hfill
\begin{minipage}{0.66\textwidth}
ionic calcium\\
cumulative distribution function\\
connection fraction\\
central nervous system\\
coefficient of variation\\
intrinsic plasticity\\
interspike interval\\
ionic potassium\\
leaky-integrate-and-fire\\
long-term depression\\
long-term potentiation\\
layer five of the mammalian neocortex\\
maximum likelihood estimation\\
neural nitric oxide synthesis\\
nitric oxide\\
probability density function\\
self-organizing recurrent neural network\\
short-term depression\\
spike-timing-dependent plasticity\\
short-term facilitation\\
short-term plasticity
\end{minipage}

\listoffigures

\listoftables

@@

Theoretical models of homeostasis are generally based on a form of feedback control. This is also known as a \textit{closed loop control} in control theory. All these systems include a number of necessary functional components, which will be briefly introduced in the following. Concepts and terms were taken from \cite[p. 59-64]{Cybernetik_Systems_Cruse_2006}.
\newpage
\subsubsection{Theory of Control Systems} \label{Control_Sys_Theo_Section}
A closed loop control generally must contain some \emph{process} that is to be regulated. Despite its potential internal complexity, its functional role in the context of a feedback system is to simply generate a time-dependent output based on the history of a time-dependent input. This output is what one wishes to control and regulate by means of the feedback loop and thus represents the signal that is further processed. It might not be possible however to directly quantify this desired output in terms of some internal state of the process. The firing rate of a neuron for example is an abstract quantity, which can not be \emph{directly} related to some neuronal state variable. Consequently, it might be necessary to transform the actual output of the process into a different signal by a \textit{feedback transducer}. The (potentially transformed) output is fed into a \textit{comparator} in the next step, which---unsurprisingly---compares the given output of the process to some given target or \textit{reference}, typically by subtraction. Note that this reference can but does note have to be a time-dependent signal. The term \textit{homeostasis} however generally refers to a fixed reference, while implementations with a time dependent reference can be regarded as a form of \textit{servomechanism}. The difference between reference and the feedback transducer's signal is known as the \textit{error} signal. This error signal is given to a \textit{controller}, which transforms this error signal into an appropriate input signal for the initially introduced process. This is also the point in the control loop where \textit{external disturbances} come into play by being added to the input generated by the controller. We illustrated the described flow of feedback control in Figure \ref{Closed_Loop_Control_Illustration}.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/illustrations/closed_loop_control.eps}
\end{center}
\caption[Schematic illustration of a closed loop control]{Schematic illustration of a closed loop control as described in Section \ref{Control_Sys_Theo_Section}}
\label{Closed_Loop_Control_Illustration}
\end{figure}
Based on the way the controller generates input for the dynamic process based on the error signal one can classify control loops by three types:
\newpage
\begin{itemize}
\item A \textit{proportional controller} generates a signal that is proportional to the error signal it receives. The most important property of a proportional controller is that the stationary state of the control loop generally does not imply an error signal equal to zero.
\item An \textit{integrator controller} uses an integration over the past of the error signal to obtain its output. The stationary solution necessarily implies that the error signal is zero.
\item A \textit{differential controller} generates its output by means of the time derivative of the error signal. This type of controller can quickly react to momentary changes but does not compensate for sustained deviations of the error signal from zero.
\end{itemize}
It should be pointed out that these types of control mechanisms can be simultaneously used in a control system, e.g. through a linear combination. Furthermore, the presented terms refer to concepts of linear control theory. In the study of nonlinear feedback systems they thus only apply in a limited regime where a first order approximation of the system is sufficiently exact.

@@

\begin{table}
\caption{Parameters of synaptic connections.}
\begin{tabu}{|l|l|l|l|l|}
\hline
\textbf{parameter} & \textbf{EE} & \textbf{EI} & \textbf{IE} & \textbf{II} \\ \hline
connection fraction & $\rightarrow 0.1$ & $0.1$ & $0.1$ & $0.5$ \\ \hline
initial connection strength & $0.0001 mV$ & $1.5 mV$ & $-1.5 mV$ & $-1.5 mV$ \\ \hline
conduction delay & $1.5 ms$ & $0.5 ms$ & $1.0 ms$ & $1.0 ms$ \\
\hline
\end{tabu}
\label{syn_conn_params}
\end{table}
\newpage

\subsubsection{Neuron Model}
We used a leaky integrate-and-fire model for all neurons in the network, whose dynamics are described by a stochastic differential equation:
\begin{equation}
{\tau_m}dV_i = -\left( V_i-E_l \right) dt + \sqrt{\tau_m} \sigma dW_i + \tau_m \sum_{j,k} w^{\rm effective}_{ij}\delta \left( t^k_{\mathrm{spike},j} + t_{\rm delay} - t \right)
\label{LIF_Dynamics}
\end{equation}
where $V_i$ is the membrane potential of neuron $i$, $E_l$ is the equilibrium membrane potential, $\tau_m$ is the time constant of the membrane, $\sigma$ is the standard deviation of the noise term and $dW$ is the standard Wiener process. A neuron is said to spike when its membrane potential reaches the threshold voltage $V_t$. The voltage is then reset to $V_r$. A refractory period was not implemented. A presynaptic spike from neuron $j$ causes a simple (delayed, see Table \ref{syn_conn_params}) increment of the membrane potential of the postsynaptic neuron $i$ by $w^{\rm effective}_{ij}$. Table \ref{LIF_neuron_params} summarizes the aforementioned set of parameters.
\begin{table}[H]
\caption{Parameters of LIF neuron}
\begin{tabu}{|l|l|l|}
\hline
\textbf{parameter} & \textbf{exc. neur.} & \textbf{inh. neur.}\\ \hline
$\mathrm{E_l}$ & $\mathrm{-60\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\tau_m}$ & $\mathrm{20\;ms}$ & $\mathrm{20\;ms}$ \\ \hline
$\mathrm{V_r}$ & $\mathrm{-70\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\sigma}$ & $\mathrm{\sqrt{5}\;mV}$ & $\mathrm{\sqrt{5}\;mV}$ \\ \hline
$\mathrm{V_t}$ & subject to IP & $\mathrm{-58\;mV}$ \\ 
\hline
\end{tabu}
\label{LIF_neuron_params}
\end{table}

@@

Finally, the dynamics of firing thresholds $V_{t,i}$ were modeled such that the rate of change is proportional to the relative deviation of $\rm NO$ concentration at the neurons' locations from a global target concentration $NO_0$.

The appropriate choice of $NO_0$ is obviously crucial for the goal of achieving and maintaining a certain level of activity. However, one cannot directly set a parameter of the model to the desired population activity as was the case for canonical intrinsic homeostasis. One rather needs to determine the average concentration \textit{associated} with the desired activity and set it as a target concentration. Though it is possible to derive this relation in an analytic fashion, we let the system run with the previous homeostatic mechanism, still solving equation \eqref{Ca_dyn}--\eqref{NO_dyn} until a steady mean over the concentrations at the neurons' positions was reached. This mean was then set to be the target concentration and we switched to diffusive homeostasis. Table \ref{Params_IP} summarizes the choice of parameters that were introduced in this section. Diffusion parameters roughly match those measured in experiments \cite{Philippides_2000}.
\begin{table}
\caption{Parameters of homeostatic intrinsic plasticity.}
\begin{tabu}{|l|l|}
\hline
\textbf{parameter} & \textbf{value} \\
\hline
$\mathrm{r_{IP}}$ & $\mathrm{3\;Hz}$ \\
\hline
$\mathrm{\eta_{IP}}$ & $\mathrm{0.1\;mV}$ \\
\hline
$\mathrm{Ca^{2+}_{spike}}$ & 1 \\ \hline
$\mathrm{\tau_{Ca^{2+}}}$ &  $\mathrm{10\;ms}$ \\
\hline
$\mathrm{\tau_{nNOS}}$ & $\mathrm{100\;ms}$ \\
\hline
$\mathrm{D}$ & default: 10 $\mathrm{\mu m^2 ms^{-1}}$ \\
\hline 
$\mathrm{\lambda}$ & $\mathrm{0.1\;s^{-1}}$ \\
\hline
$\mathrm{\tau_{V_t}}$ & see Section \ref{activ_analys} \\
\hline
\end{tabu}
\label{Params_IP}
\end{table}

@@

3.) Dirichlet boundary conditions with $NO = NO_{\rm bound.}$ at the boundaries.\\
Neumann boundary conditions were used for most of the simulations if not explicitly marked differently. This decision relates to the previously described mechanism of synaptic growth: Neurons placed close to the edge of the sheet have a lower connection probability due to the absence of neighboring neurons in the direction perpendicular to the close-by border. It therefore models the synaptic growth within a square ``cutout" of neural tissue. The Neumann boundary condition fits into this picture since it allows a zero-flux condition at the borders. This is a reasonable assumption, because NO molecules cannot diffuse out of the tissue (unless they were placed in a fluid surrounding).

@@

\subsection{Activity Analysis} \label{activ_analys}
As a first step, we investigated how diffusive homeostasis compares to non-diffusive homeostasis in terms of maintaining a stable level of population activity among excitatory and inhibitory population (though the latter one is only indirectly affected by excitatory synaptic input).
\subsubsection{Dynamic Behavior of the Population Activity}
We set the time constant of threshold adaptation in the diffusive homeostasis to $\mathrm{2500\, ms}$ on a first attempt, as given by \cite{Sweeney_Paper}. Switching between non-diffusive/diffusive homeostasis after $\rm 750\, s$, Figure \ref{full_sim_osci} shows the resulting dynamics of the population activity of the excitatory and inhibitory group (panel A/A*), the average $\rm NO$ concentration at the excitatory neurons' positions (panel B/B*) and the average firing threshold (panel C/C*) within the excitatory group (inhibitory threshold was fixed, see table \ref{LIF_neuron_params}). Both homeostatic mechanisms managed to keep the excitatory population activity in the desired range of 3 Hz. However, what might appear to be slightly faster and stronger random fluctuations in panel A--C were in fact regular oscillations across all three depicted variables, as can be seen in panel A*--C*, showing the dynamics in a smaller time window. While the oscillation amplitudes undergo a rather unpredictable course, the frequency remains at a constant level of $\simeq 0.5\, Hz$. This feature is also illustrated by the fact that the power spectra depicted in Figure \ref{Power_Spec_without_Analysis} have a peak at the frequency that appeared to be dominant in Figure \ref{full_sim_osci}. The fact that the mean over all power spectra of excitatory thresholds differs from the power spectrum of the mean of these thresholds in its amplitude indicates that not all thresholds oscillate at the same phase. The overall shapes of both spectra are still similar, though.
\begin{figure}
\begin{center}
%% oscillating_activity_plot.py
\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_compl.png}
\end{center}
%\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_closeup.png}
\caption[Time course of population firing rates, population mean of NO concentration and of excitatory threshold]{A/A*: Mean of excitatory (blue) and inhibitory (red) neuronal firing rate. B/B*: NO concentration averaged over readouts at excitatory neurons' positions. C/C*: Average excitatory firing threshold. Non-diffusive homeostasis was used for $\mathrm{0-700\, s}$ , diffusive h. for $\mathrm{700-1500\, s}$. Target activity was $\mathrm{3\, Hz}$.}
\label{full_sim_osci}
\end{figure}
\begin{figure}
\begin{center}
%%% plot_pow_spec.py
\includegraphics[width=0.7\textwidth]{../../plots/power_spec/power_spectrum_sim_only.png}
\caption[Power spectrum of thresholds of excitatory neurons]{Blue: Mean of power spectrum over all excitatory thresholds. Red: Power spectrum of mean over all excitatory thresholds.}
\label{Power_Spec_without_Analysis}
\end{center}
\end{figure}

@@

where $\tau_e$ and $\tau_i$ are time constants representing the rapidity of firing rate dynamics, $\phi_e$ and $\phi_i$ the respective transfer functions between synaptic input and firing rate and $W_{xy}$ the mean synaptic weight from population x to population y. A persistent oscillation then corresponds to a stable limit cycle in the two-dimensional phase plane. Such a limit cycle exists inside a region $R$ if $R$ contains no fixed points and if any trajectory whose starting points lies within $R$ is confined to $R$, according to the Poincaré-Bendixson theorem. The second condition is equivalent to the condition that for all points $\mathbf{r}_{\rm edge}$ on the edge of $R$, the vector field $\dot{\mathbf{r}}_{\rm edge}$ is facing into $R$, see \cite[p.~248]{Dyn_Sys_Hirsch}. This case occurs if the system has a locally unstable fixed point, but a non-linearity that prevents an infinite deviation from the fixed point.

@@

\caption[Comparison of mean firing rate of a LIF neuron without noise and with noise]{Comparison of mean firing rate of a LIF neuron without noise (squares) and with noise (circles, $\mathrm{\sigma = \sqrt{5}\, mV}$), averaged over $\mathrm{50\, s}$. Red and blue curves are the corresponding analytic predictions, see equation \ref{LIF_Fir_Rate_no_Noise} and \ref{LIF_Fir_Rate_with_Noise1}.}

@@

\sigma_{\rm tot.} = \sqrt{\sigma_{\rm intr.}^2 + \sum_A \sigma_{AB}^2} \; .

@@

x_{\rm st.}(r) &= \frac{1-\exp \left(-\frac{1}{r\cdot \tau _d}\right)}{1-\left(1-u_{st.}\left(r\right)\right)\cdot \exp \left(-\frac{1}{r\cdot \tau _d}\right)} \label{STP_steady1} \\
u_{\rm st.}(r) &= \frac{U}{1-\left(1-U\right)\exp \left(-\frac{1}{r\cdot \tau _f}\right)} \; .\label{STP_steady2}

@@

One thus needs to incorporate the factor $x_{\rm st.}\cdot u_{\rm st.}$ by means of the previously described mean input and the contribution to the total variance of the membrane noise when trying to find the fixed point of Equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}.

All these aspects taken together make it impossible to find an analytic expression for the fixed point of Equation \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}. Running a simulation of Equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}, including all aforementioned approximations resulted in a stable configuration. We set the relevant parameters to their predefined values for this simulation and additionally extracted the mean threshold of the excitatory population which was set by the non-diffusive homeostasis, see C/C* in Figure \ref{full_sim_osci}, yielding a threshold of $\mathrm{-56.963\, mV}$. The results are depicted in Figure \ref{Dyn_Rate_Approx}. The system settled to a fixed point of $\mathrm{(r_{0,e},r_{0,i}) = (2.992\, Hz, 7.144\, Hz)}$, which is quite close to $\mathrm{(3\, Hz, 6.768\, Hz)}$, being the mean excitatory and inhibitory firing rate found between $\mathrm{100-750\, s}$ of the full simulation (i.e., during non-diffusive homeostasis).
\begin{figure}
%% fixed_point_net.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/approx_rates_EI.png}
\end{center}
\caption[Dynamics of rate-based network model]{Dynamics of equations \ref{2d-exc_inh_rate_1} (blue line) and \ref{2d-exc_inh_rate_2} (red line). Dotted lines mark the mean frequencies that were present in the full spiking network. Excitatory: 3 Hz, Inhibitory: 6,768 Hz. \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} converged to 2.992 Hz and 7.144 Hz respectively.}
\label{Dyn_Rate_Approx}
\end{figure}
In principle, one could now further quantify the stability of the found fixed point by calculating the Jacobian matrix at $(r_{0,e},r_{0,i})$. For the sake of the initial question, though, namely the source of oscillation, it is sufficient at this point to restrict oneself to a preliminary result: The recurrent network as such (without homeostasis) is stable under the given choice of parameters, assuming validity of the described rate model. Furthermore, on an empirical basis, the occurrence of oscillations appears to be rather dependent on the choice of homeostasis since all parameters of the simulation were kept unchanged upon the transition after 750 s, apart from the homeostatic mechanism itself. Therefore, we present in the following section our further analysis of the dynamics underlying the diffusive homeostatic feedback.

@@

Both forms of homeostasis use excitatory firing thresholds as a means of adjusting the excitatory firing rate. Though having a relatively immediate impact on the firing rate of the particular neuron, the network as a whole reacts by means of a local disturbance of activity as well: It settles at a new fixed point of firing rates. Two questions are of importance in the context of feedback dynamics. First---obviously---how the relation between a local change of threshold and the new fixed point of the network can be described and second, whether the time scale of the network's response is of relevance with respect to other time scales involved in the feedback loop. Regarding the second question, Figure \ref{rate_th_close} suggests that the excitatory population activity follows the mean firing threshold in a quasi instantaneous fashion, at least in comparison with the overall time scale of the oscillating pattern.
\begin{figure}
%% oscillating_activity_plot_rate_threshold.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/diff_hom/rate_th_close.png}
\end{center}
\caption[Time course of the mean firing rate and the mean firing threshold within the excitatory pool]{Time course of the mean firing rate (blue) and the mean firing threshold within the excitatory pool (red). Bin width for firing rate estimation was 0.1 s.}
\label{rate_th_close}
\end{figure}
We thus presumed an immediate functional relationship $\mathbf{r}_e (t) = \mathbf{r}_e (\mathbf{V}_{t,e} (t))$ in a first attempt at understanding the occurrence of undesired oscillations, where $\mathbf{r}_e$ and $\mathbf{V}_{t,e}$ represent the set of excitatory firing rates and thresholds, respectively.

@@

\caption[Distribution of firing rates, full simulation and analytic prediction]{Distribution of firing rates for $\mathrm{D=10\, \mu m^2 ms^{-1}}$, full simulation and analytic prediction. Data was taken from 10 simulation runs.}
